Energy Clustering

### 2017-10-08 energy_clustering/draft/energy_clustering.tex
more comments

% @gui: have other people shown a relationship between kernel k-means and 
% say, RKHS, or something else? 
% @jovo: any kernel method is defined on a RKHS. 
% For instance ref [11] is the first one to consider kernel k-means in a 
% RKHS. The problem is how do you formulate a 
% statistical problem in a RKHS, or how do you know if there is a kernel
% related to it. In the case of energy statistics this was done by [6], and
% is a nontrivial result. It's based on this work that we have the theory 
% presented in this paper

@@

% @jovo: I corrected pretty much all your previous points, and also re-wrote 
% a lot of other stuff throughout the paper. 
% I deleted your previous comments which I believe were taken
% care of, since most of them were simple. The ones which were not, I will
% leave it, in the beggining of each section or in the middle.
%
% @jovo: About ToC, I agree and wouldn't mind remove it, however it is a 
% tradition that most referees like to see. In this case I also included 
% some important information about our work, and it helps to make a transition 
% between introduction and the remaining of the paper. 
% I would leave it, but if you prefer to remove it, let me know and I 
% will adapt the paper.

@@

% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also
% said to be non-parametric. I think non-parametric tests also mean that
% it does not assume any form of distributions, i.e. it is a distribution-free
% statistical test.

@@

% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
% @jovo: ok, will do that after we finish everything. There is a package that
% automatically remove numbers from unreferenced equations. It is a common
% practice however to number every equation in the paper,
% specially for reviewing purposes. But later 
% I will remove the numbers from the not cited ones if you don't like

% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
% @jovo: I will do that for the main symbols, and refer to equations only
% when necessary. But sometimes it is unnavoidable, specially
% when referring to long 
% expressions or derivation steps. It is way more clear and precise to say, 
% replacing equation (1) into (2) we obtain ... than saying, 
% replacing A into B we obtain ... and the reader has to look around
% the whole paper to figure it out what  A, B are

% @gui: do other people know this result? (about min W = max X) if so, cite them, if not, state that.
% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
% in their paper, however, they didn't write a proof. They should have
% it in some other paper.  However, they mention 
% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
% here I did for a general semimetric. Thus I don't really know if in general
% this is equality is known or not, but it should because it's a very easy
% derivation. The PhD thesis do not 
% mention anywhere that minimizing W is equivalent to maximizing S.
% Therefore, I don't want to sound like this is a big result and possibly
% a referee can say  "well this is trivial and known".
% Thus I just preferred to do my own derivation and not say anything.
% What do you think??

% @gui: i don't understand the notaiton (.)_jj. is a sum missing?
% @jovo: this is standard.  A_{ij} denotes the 
% $(i,j)$th element of matrix A as you know. 
% Since here A = Z^T G Z is a product of 3 matrices
% it is common to just put a parenthesis to avoid confusion, that's all

% @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
% @jovo: This was a long issue this last week as you know. 
% The spectral initialization is, often, very
% close to a local optimum of E-clustering algorithm, and kernel k-means
% as well. Now, since the set of local optima of Hartigan's is a subset of 
% those of Lloyd's, if you initialize "too well", you might get stuck on 
% a not so great local optimum of Lloyd's, which otherwise Hartigan's 
% would be able to escape. Numerically, in most cases Hartigan's does
% better when initializing with k-means++, or even at random.

% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
% @jovo: I tried to not commit to a single view. We show they are
% equivalent, when opperating on the same kernel. However, kernel k-means per
% se does not consider anything about the kernel choice. Energy statistics,
% however, provides a theory valid on spaces of negative type, which guarantees
% that a kernel constructed within this formalism will be 
% positive-semidefinite. Thus, I still prefer to look in the first direction.

@@

\begin{proposition} 
\label{th:qcqp2}
The optimization problem \eqref{eq:minimize} is equivalent to
\begin{equation}
\label{eq:qcqp2}
\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right)  \qquad
\mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
$\Zt \Zt^\top \e = \e$},
\end{equation}
where $\e = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
and $G$ is the Gram matrix \eqref{eq:kernel_matrix}.
\end{proposition}
\begin{proof}
From 
\eqref{eq:gen_kernel},
\eqref{eq:g_def}, and
\eqref{eq:within}
we have
\begin{equation}
\label{eq:W2}
W(\C_1,\dotsc,\C_k  )
= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg).
\end{equation}
Note that the first term is global so it does not contribute to the 
optimization problem.
Therefore, minimizing \eqref{eq:W2} is equivalent to
\begin{equation}
\label{eq:max_prob}
\max_{ \C_1,\dotsc,\C_k } 
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
\end{equation}
But 
\begin{equation}
\sum_{x, y \in \C_j} \kk(x, y) =
\sum_{p=1}^{n} \sum_{q=1}^{n} Z_{pj} Z_{qj} G_{pq} = 
(Z^\top G \, Z)_{jj},
\end{equation}
where we used the definitions \eqref{eq:kernel_matrix} and
\eqref{eq:label_matrix}. Thus, the objective function in 
\eqref{eq:max_prob} is equal to
$\Tr \left( D^{-1} Z^\top G Z \right)$. 
% @gui: the above step is not obvious
% @jovo: It is obvious. 
% From (21) we have \sum_j 1/n_j (Z^T G Z)_jj. But 1/n_j is just
% D^{-1}_jj (the diagonal matrix). The sum over j goes over the diagonals of
% D^{-1} Z^T G Z which is the definition of the trace
Now we can
use the cyclic property
of the trace, and by the  definition of the matrix $Z$
in \eqref{eq:label_matrix}, we obtain the following integer
programing problem:
\begin{equation}\label{eq:qcqp}
\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
\big( ZD^{-1/2} \big) 
\Big) \quad
\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
$\sum_{i=1}^n Z_{ij} = n_j$.}
\end{equation}

@@

Consider the ratio association problem \eqref{eq:assoc}, 
whose objective function can be written as
\begin{equation}
\sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
\mathcal{A}_{pq} = \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
\dfrac{Z_{qj}}{\sqrt{s_j}}
= \Tr\left( Y^\top \mathcal{A} Y \right) ,
\end{equation}
with $Z$ defined in \eqref{eq:label_matrix} and $Y$ in
\eqref{eq:weighted_matrices}. Therefore, the ratio association problem
can be written in the form \eqref{eq:qcqp3} as
\begin{equation}
\max_H \Tr\left( H^\top \mathcal{W}^{-1/2} \mathcal{A} \mathcal{W}^{-1/2} H 
\right) \qquad \mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
\om=\om$}.
\end{equation}
This is exactly the same problem as weighted energy 
clustering with 
$G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming this
matrix is positive semidefinite, this generates a semimetric
\eqref{eq:gen_kernel} for graphs given by
\begin{equation}
\label{eq:metric_graphs}
\rho(i,j) = 
\dfrac{\mathcal{A}_{ii}}{w_i^{2}}
+\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
-\dfrac{2 \mathcal{A}_{ij}}{w_i w_j} \qquad\mbox{or}\qquad
\rho(i,j) = -\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
\end{equation}
for vertices $i,j \in \mathcal{V}$, and 
where in the second equation we assume the graph has no self-loops,
i.e. $\mathcal{A}_{ii} = 0$. 
Using \eqref{eq:metric_graphs} 
in the energy statistics formulation 
allows one to make inference
on graphs.
% @gui: can you provide some guidance here, in particular, what do the weights "mean" or "correspond to"? right now, it feels like just math.
% @jovo: ok, added the phrase below
% let me know if not enough. In practice the weights can be anything attributed
% to the nodes of the graph
Above, the weight $w_i = w(i)$ of node $i\in \mathcal{V}$ can be, for instance,
its degree $w_i = d(i)$.

@@

% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
% @jovo: it does not require initialization in the sense that we do not
% assign points to clusters as an initial step, as k-means++ for instance, or
% random initialization. This algorithm is "exact" since nothing
% is random in it, and multiple runs will always give the same answer

% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.  i'm not sure the best way to visualize it, maybe play with a few options.
% @jovo: I didn't do this yet and actually not sure how to do it.
% Are you thinking about this: 
% https://en.wikipedia.org/wiki/Kernel_density_estimation
% Also, I'm not sure if the kernel from energy integrates to 1, etc. and is
% this really worthy it?

% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
% @jovo: yes, and GMM would do great! In 1D is very hard to find settings where
% GMM and k-means perform poorly, so we have to cherry pick a little. In 1D you
% basically do not have much choice, if data has minimal separation k-means and
% GMM do well. The purpose is to
% illustrate that there are settings where k-means/GMM do poorly and energy
% does not. If I choose the same parameters as before all algorithms are doing
% great, sometimes GMM can even do better than energy. All I did was to
% translate one of the lognormals mu2=-1.5 and give it a big variance, so both
% distributions overlap a lot, but the overlaping is essentially on the tail 
% of one of them

@@

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-.8em](b)
\end{minipage}
\caption{
\label{fig:1d}
$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM in one dimension.
The two clusters have the same number of points, and we use
$50$ Monte Carlo runs.
We plot the average 
accuracy \eqref{eq:accuracy} versus the total number of points, 
where error bars are standard error. The dashed line indicates 
Bayes accuracy. (a) Data distributed as \eqref{eq:two_normal}, where the
optimal accuracy is $\approx 0.956$.
(b) Data from \eqref{eq:two_lognormal} where the optimal
accuracy is $\approx 0.852$.
}
\end{figure}

@@

% @jovo: check this paragraph. I included it to answer your question
% about having K(x,x) and \ell=1,...,k without exluding \ell=j. Essentially,
% the term K(x,x) will be hidden inside the mean of the cluster to which
% x is currently assigned to. This also clarifies the connection to 
% the standard formulation of kernel k-means.
Obviously, $\mathcal{E}^{L}$-clustering is precisely
kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
but written in a more concise and clear manner. Indeed, recalling that
$\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
\to \mathcal{H}_\kk$ is the feature map, we have from \eqref{eq:Jell} that
\begin{equation}
%\begin{split}
J^{(\ell)}(x_i) 
%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
= 
\langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
-2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
= 
\| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
%\end{split}
\end{equation}
where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
is the mean of cluster $\C_\ell$. 
Therefore, $\min_\ell J^{(\ell)}(x_i) = 
\min_\ell \| \varphi(x_i) - \varphi(\mu_\ell)\|^2$, i.e. we are assigning
$x_i$ to the cluster with closest center (in feature space),
which is the familiar Lloyd's
heuristic approach that kernel $k$-means is based upon.

@@

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, initial label
    matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
        have the number of points in each cluster%, obtained 
        %from $D = Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$,
            where $J^{(\ell)}(x_i)$ is defined in \eqref{eq:Jell}
            %, for $\ell=1,2,\dots,k$
        \IF{ $j^\star \ne j$} 
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
$\mathcal{E}^{L}$-clustering is Lloyd's method for energy clustering, which
is precisely~~~~~~~~ \\ 
kernel $k$-means algorithm. This procedure finds
local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

We thus propose an algorithm that goes as follows.
We start with an initial configuration for the label matrix $Z$, 
then for each
point $x_i$ 
we compute the cost of moving it to another cluster $\C_\ell$, i.e.
$\Delta Q^{j\to \ell}(x_i)$ for 
$\ell=1,\dots,k$ with $\ell \ne j$, where $j$ denotes the index of its current
partition, $x \in \C_j$. Hence, we choose
\begin{equation}
j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
\Delta^{j \to \ell}(x_i).
\end{equation}
If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
we keep $x_i$ in its original cluster $\C_j$. 
This process is repeated
until no points are assigned to new clusters. 
The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
coin $\mathcal{E}^H$-clustering to emphasize that it is based on
Hartigan's method.
This method automatically ensures that the objective function is
monotonically increasing at each iteration, and consequently the algorithm
converges in a finite number of steps.

>>>>

% @gui: you didn't update the abstract?

@@

% @jovo: About ToC, I agree and wouldn't mind remove it, however it is a 
% tradition that most referees like to see. In this case I also included 
% some important information about our work, and it helps to make a transition 
% between introduction and the remaining of the paper. 
% I would leave it, but if you prefer to remove it, let me know and I 
% will adapt the paper.

% @gui: my experience is that most reviewers absolutely do *NOT* want it, rather, we skip it, and journals often remove it prior to publication.

@@

% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
% @jovo: I stopped saying non-parametric, however, Kolmogorov-Smirnov is also
% said to be non-parametric. I think non-parametric tests also mean that
% it does not assume any form of distributions, i.e. it is a distribution-free
% statistical test.
% @gui: KS test is nonparametric because it can reject for any distribution, we cannot cluster for any distribution, so we are not.

@@

% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
% @jovo: ok, will do that after we finish everything. There is a package that
% automatically remove numbers from unreferenced equations. It is a common
% practice however to number every equation in the paper,
% specially for reviewing purposes. But later 
% I will remove the numbers from the not cited ones if you don't like
% @gui: maybe common in math, but certainly not in stats / data science. 
% in any case, what is the package that does that?

% @gui: do other people know this result? (about min W = max X) if so, cite them, if not, state that.
% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
% in their paper, however, they didn't write a proof. They should have
% it in some other paper.  However, they mention 
% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
% here I did for a general semimetric. Thus I don't really know if in general
% this is equality is known or not, but it should because it's a very easy
% derivation. The PhD thesis do not 
% mention anywhere that minimizing W is equivalent to maximizing S.
% Therefore, I don't want to sound like this is a big result and possibly
% a referee can say  "well this is trivial and known".
% Thus I just preferred to do my own derivation and not say anything.
% What do you think??
% @gui: i'd state it as a lemma then.

% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
% @jovo: I tried to not commit to a single view. We show they are
% equivalent, when opperating on the same kernel. However, kernel k-means per
% se does not consider anything about the kernel choice. Energy statistics,
% however, provides a theory valid on spaces of negative type, which guarantees
% that a kernel constructed within this formalism will be 
% positive-semidefinite. Thus, I still prefer to look in the first direction.
% @gui: ok, did you update the text? i don't recall where i made this comment and don't want to search for it again. can you point me to the place that you say this?

@@

\begin{proposition} 
\label{th:qcqp2}
The optimization problem \eqref{eq:minimize} is equivalent to
\begin{equation}
\label{eq:qcqp2}
\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right)  \qquad
\mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
$\Zt \Zt^\top \e = \e$},
\end{equation}
where $\e = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
and $G$ is the Gram matrix \eqref{eq:kernel_matrix}.
\end{proposition}
\begin{proof}
From 
\eqref{eq:gen_kernel},
\eqref{eq:g_def}, and
\eqref{eq:within}
we have
\begin{equation}
\label{eq:W2}
W(\C_1,\dotsc,\C_k  )
= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg).
\end{equation}
Note that the first term is global so it does not contribute to the 
optimization problem.
Therefore, minimizing \eqref{eq:W2} is equivalent to
\begin{equation}
\label{eq:max_prob}
\max_{ \C_1,\dotsc,\C_k } 
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
\end{equation}
But 
\begin{equation}
\sum_{x, y \in \C_j} \kk(x, y) =
\sum_{p=1}^{n} \sum_{q=1}^{n} Z_{pj} Z_{qj} G_{pq} = 
(Z^\top G \, Z)_{jj},
\end{equation}
where we used the definitions \eqref{eq:kernel_matrix} and
\eqref{eq:label_matrix}. Thus, the objective function in 
\eqref{eq:max_prob} is equal to
$\Tr \left( D^{-1} Z^\top G Z \right)$. 
% @gui: the above step is not obvious
% @jovo: It is obvious. 
% From (21) we have \sum_j 1/n_j (Z^T G Z)_jj. But 1/n_j is just
% D^{-1}_jj (the diagonal matrix). The sum over j goes over the diagonals of
% D^{-1} Z^T G Z which is the definition of the trace
% @gui: obvious to you \neq obvious.  i understood, i want you to explain it.
Now we can
use the cyclic property
of the trace, and by the  definition of the matrix $Z$
in \eqref{eq:label_matrix}, we obtain the following integer
programing problem:
\begin{equation}\label{eq:qcqp}
\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
\big( ZD^{-1/2} \big) 
\Big) \quad
\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
$\sum_{i=1}^n Z_{ij} = n_j$.}
\end{equation}

@@

Consider the ratio association problem \eqref{eq:assoc}, 
whose objective function can be written as
\begin{equation}
\sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
\mathcal{A}_{pq} = \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
\dfrac{Z_{qj}}{\sqrt{s_j}}
= \Tr\left( Y^\top \mathcal{A} Y \right) ,
\end{equation}
with $Z$ defined in \eqref{eq:label_matrix} and $Y$ in
\eqref{eq:weighted_matrices}. Therefore, the ratio association problem
can be written in the form \eqref{eq:qcqp3} as
\begin{equation}
\max_H \Tr\left( H^\top \mathcal{W}^{-1/2} \mathcal{A} \mathcal{W}^{-1/2} H 
\right) \qquad \mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
\om=\om$}.
\end{equation}
This is exactly the same problem as weighted energy 
clustering with 
$G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming this
matrix is positive semidefinite, this generates a semimetric
\eqref{eq:gen_kernel} for graphs given by
\begin{equation}
\label{eq:metric_graphs}
\rho(i,j) = 
\dfrac{\mathcal{A}_{ii}}{w_i^{2}}
+\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
-\dfrac{2 \mathcal{A}_{ij}}{w_i w_j} \qquad\mbox{or}\qquad
\rho(i,j) = -\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
\end{equation}
for vertices $i,j \in \mathcal{V}$, and 
where in the second equation we assume the graph has no self-loops,
i.e. $\mathcal{A}_{ii} = 0$. 
Using \eqref{eq:metric_graphs} 
in the energy statistics formulation 
allows one to make inference
on graphs.
Above, the weight $w_i = w(i)$ of node $i\in \mathcal{V}$ can be, for instance,
its degree $w_i = d(i)$.

@@

% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
% @jovo: it does not require initialization in the sense that we do not
% assign points to clusters as an initial step, as k-means++ for instance, or
% random initialization. This algorithm is "exact" since nothing
% is random in it, and multiple runs will always give the same answer
% @gui: i know what you mean, my point is that it is unclear.  state in the manuscript what you mean: "it is a deterministic algorithm." or state something else, that it does not require an initialization i don't feel is correct or informative.

% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.  i'm not sure the best way to visualize it, maybe play with a few options.
% @jovo: I didn't do this yet and actually not sure how to do it.
% Are you thinking about this: 
% https://en.wikipedia.org/wiki/Kernel_density_estimation
% Also, I'm not sure if the kernel from energy integrates to 1, etc. and is
% this really worthy it?
% @gui: yes, it is really worth it, that is why i suggested it. it is trivial to do.  currently, the reader has no idea *why* the different algs get different answers.

% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
% @jovo: yes, and GMM would do great! In 1D is very hard to find settings where
% GMM and k-means perform poorly, so we have to cherry pick a little. In 1D you
% basically do not have much choice, if data has minimal separation k-means and
% GMM do well. The purpose is to
% illustrate that there are settings where k-means/GMM do poorly and energy
% does not. If I choose the same parameters as before all algorithms are doing
% great, sometimes GMM can even do better than energy. All I did was to
% translate one of the lognormals mu2=-1.5 and give it a big variance, so both
% distributions overlap a lot, but the overlaping is essentially on the tail 
% of one of them
% @gui: ok, well why not use the same params from fig 1b in fig 1a then?

@@

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-.8em](b)
\end{minipage}
\caption{
\label{fig:1d}
$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM in one dimension.
The two clusters have the same number of points, and we use
$50$ Monte Carlo runs.
We plot the average 
accuracy \eqref{eq:accuracy} versus the total number of points, 
where error bars are standard error. The dashed line indicates 
Bayes accuracy. (a) Data distributed as \eqref{eq:two_normal}, where the
optimal accuracy is $\approx 0.956$.
(b) Data from \eqref{eq:two_lognormal} where the optimal
accuracy is $\approx 0.852$.
% @gui: if you insist on using the equations, at least also mention something semantic about them, ie, Data distributed Gaussian (as in 46), or Data distributed log Gaussian (as in 47).  figures & captions should stand on their own without requiring referring to the paper.
}
% @gui: why did you take away the plot of the distribution? those are incredibly helpful.  
\end{figure}

@@

% @jovo: check this paragraph. I included it to answer your question
% about having K(x,x) and \ell=1,...,k without exluding \ell=j. Essentially,
% the term K(x,x) will be hidden inside the mean of the cluster to which
% x is currently assigned to. This also clarifies the connection to 
% the standard formulation of kernel k-means.
Obviously, $\mathcal{E}^{L}$-clustering is precisely
kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
but written more concisely and with the kernel induced by energy statistics. Indeed, recalling that
$\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
\to \mathcal{H}_\kk$ is the feature map, we have from \eqref{eq:Jell} that
\begin{equation}
%\begin{split}
J^{(\ell)}(x_i) 
%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
= 
\langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
-2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
= 
\| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
%\end{split}
\end{equation}
where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
is the mean of cluster $\C_\ell$. 
Therefore, $\min_\ell J^{(\ell)}(x_i) = 
\min_\ell \| \varphi(x_i) - \varphi(\mu_\ell)\|^2$, i.e. we are assigning
$x_i$ to the cluster with closest center (in feature space),
which is the familiar Lloyd's
heuristic approach that kernel $k$-means is based upon.

@@

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, initial label
    matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
        have the number of points in each cluster%, obtained 
        %from $D = Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$,
            where $J^{(\ell)}(x_i)$ is defined in \eqref{eq:Jell}
            %, for $\ell=1,2,\dots,k$
        \IF{ $j^\star \ne j$} 
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
$\mathcal{E}^{L}$-clustering is Lloyd's method for energy clustering, which
is precisely~~~~~~~~ \\ 
kernel $k$-means algorithm, with the kernel induced by energy statistics. This procedure finds
local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
% @gui: spacing is weird.
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

We thus propose an algorithm that goes as follows.
We start with an initial configuration for the label matrix $Z$, 
then for each
point $x_i$ 
we compute the cost of moving it to another cluster $\C_\ell$, i.e.
$\Delta Q^{j\to \ell}(x_i)$ for 
$\ell=1,\dots,k$ with $\ell \ne j$, where $j$ denotes the index of its current
partition, $x \in \C_j$. Hence, we choose
\begin{equation}
j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
\Delta^{j \to \ell}(x_i).
\end{equation}
If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
we keep $x_i$ in its original cluster $\C_j$. 
This process is repeated
until no points are assigned to new clusters. 
The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
denote $\mathcal{E}^H$-clustering to emphasize that it is based on
Hartigan's method.
This method automatically ensures that the objective function is
monotonically increasing at each iteration, and consequently the algorithm
converges in a finite number of steps.

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, 
                initial label matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the energy costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
        in each cluster%, obtained from $D=Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
            %for $\ell=1,2,\dots,k$ and $\ell \ne j$; 
            using \eqref{eq:changeQ} \label{stepmove}
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and 
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ G_{ii}$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
$\mathcal{E}^H$-clustering is Hartigan's method for energy clustering.
This algorithm~~~~~~ \\ finds local solutions to  
the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
% @gui: clarify precisely which steps are different from alg 3 in caption.
\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}
